---
title: "Workshop 2: Advanced data wrangling: Extracting ecological signals from noisy systems"
author: "Celine M. Egsaeter"
format:
  html:
    embed-resources: true
execute:
  warning: false
  message: false
---

#### Housekeeping
```{r}
## Inventory every active object currently residing in session RAM
objects()
## Purge global environment
rm(list = ls())
# Confirm that global session memory now completely vacant
objects()
```

#### Library
```{r}
# Load the primary data science framework
library(tidyverse)
library(palmerpenguins)
library(lubridate)
library(readxl) # For reading in excel files
```


# 2.2 Tidying data

1.	Put each dataset in a tibble (special dataframe)

**table 1**
```{r}
#Creating tibble from the data in the workbook
table1 <- tibble(
  country = c("Afghanistan", "Afghanistan", "Brazil", "Brazil", "China", "China"),
  year = c(1999, 2000, 1999, 2000, 1999, 2000),
  cases = c(745, 2666, 37737, 80488, 212258, 213766),
  population = c(19987071, 20595360, 172006362, 174504898, 1272915272, 1280428583)
) # This is a tidy datasett
table1
```

**table 2**
```{r}
table2 <- tibble(
  country = c("Afghanistan", "Afghanistan","Afghanistan","Afghanistan", "Brazil", "Brazil", "Brazil", "Brazil", "China", "China","China", "China"),
year = c(1999,1999,2000, 2000,1999,1999,2000, 2000,1999,1999,2000, 2000),
type =  c("cases","population","cases","population", "cases","population","cases","population","cases","population","cases","population"),
count = c(745, 19987071, 2666,20595360,37737,172006362,80488,174504898,212258,1272915272,213766,1280428583)) # This data is the same but is messy
table2
```

**table 3**
```{r} 
table3 <- tibble(
  country = c("Afghanistan", "Afghanistan", "Brazil", "Brazil", "China", "China"),
  year = c(1999, 2000, 1999, 2000, 1999, 2000),
  rate = c("745/19987071", "2666/20595360", "37737/172006362",
           "80488/174504898", "212258/1272915272", "213766/1280428583")
)
table3
```


```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases)
```


```{r}
# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

```

# 2.3 Pivoting data to make it tidy

# 2.4 Lengthening datasets

```{r}
# We are using the data billboard as practise
billboard
```
**pivot longer**
```{r}
billboard %>% pivot_longer( # Pivot longer creates a long df of a wide one.
  cols = starts_with("wk"), # chooses the collums starting with wk
  names_to = "week", # puts the names from wk, into a collumn named week
  values_to = "rank", # adds the values to a coloumn named rank
  values_drop_na = TRUE
)
```

**Pivoting longer - example**
```{r}
df <- tribble( # Save the tribble as a dataset called df
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)
df
```

However, we want our new (tidy) dataset to have three variables: 
1.	id (which already exists)
2.	measurement (the column names) 
3.	value (the cell values)

**Pivot longer**
we use pivot longer to get 3 variables
```{r}
df %>% 
  pivot_longer(
    cols = bp1:bp2, # The chosen columns to that we are going to pivot into rows 
    names_to = "measurement", # The names is the bp1 and bp2 theire collumn is going to be called measurment 
    values_to = "value" # The values they hold will go into a difrent collumn called value
  )

```

# 2.5 Widening datasets

Lengthening the dataset is the oposite of widening it, and we use the function pivot_wider

For this i will use an example from R4DS, the cms_patient_experience dataset 

```{r}
cms_patient_experience %>% 
  distinct(measure_cd, measure_title) # shows the complete sett of values of the df
```

**Pivot wider**
```{r}
cms_patient_experience %>% 
    pivot_wider(
    id_cols = starts_with("org"), # Says which columns identify a row, and pics every column that starts with "org"
    names_from = measure_cd, # says whcih column values become the new column names
    values_from = prf_rate # says which column supplies the numbers that fill the new columns.
  )
```

Adding the "id_cols = starts_with("org")" line helped us make the output we wanted, not adding this line made the df output look a bit off. 

**Pivoting wider - example**
Changed df to df_2 so it is not the same as the earlier example.

```{r}
# Creating the df
df_2 <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```


```{r}
df_2 %>% 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```

```{r}
df_2 %>% 
  distinct(measurement) |> 
  pull() #chek the unique values
```
```{r}
df_2 %>% 
  select(-measurement, -value) |> 
  distinct()
```

pivot_wider can now combie these results to generate an empty df

```{r}
df_2 %>% 
  select(-measurement, -value) %>% 
  distinct() %>%  
  mutate(x = NA, y = NA, z = NA) 
# creates a empty df
```


# 2.6 Pivoting Exercises using Palmer Penguins

**Lengthening data with** pivot_longer()

```{r}
#Create a long df version of the penguins df
penguins_long <- penguins %>% 
  pivot_longer(
    cols = c(bill_length_mm,bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
# view the result
head(penguins_long) 
```

**Create ggplot**
```{r}
penguins_long %>% 
  drop_na(value) %>% # Remover the NA values in the value column
    ggplot(aes(x = value, fill = species)) + # Tells ggplot what values go where
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") + 
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )

```

**Widening data with** pivot_wider()
Time to do the oposite

```{r}
# Calculate the mean body mass for each species
mass_summary <- penguins %>% 
  drop_na(body_mass_g) %>% 
  group_by(species, island) %>% 
  summarise(mean_mass = mean(body_mass_g))
# View the df
head(mass_summary)

```
The above verion is tidy but not great for publication

```{r}
# Creating a clean matrix of species by Isalnd
mass_matrix <- mass_summary %>% 
  pivot_wider( #Using pivot_wider to crate a column for each island
    names_from = island,
    values_from = mean_mass
  )
# view the df
head(mass_matrix)

```


# 2.7 Separating and uniting data tables

Using the function separete() to tidy table 3 from earlier

**Here’s table3:**
```{r}
head(table3)
```

I want to split the rate column up into two variables 
1. cases 
2. Population

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))

# same code but uses the "sep=" to specify what charakter to separate into two coloumns

table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")

# In both these something went wrong, i am asuming it is beceause both tibbles have population and cases in the chr class
```

**Note** Problem above i needed to change the way i created the tibble at the begining, the rate had to be written in quatas, sutch as "745/19987071", if not R does the calculation and removes the "/". 

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE) # Convets = TRUE keeps the two new columns as int

```

**Using** unite():
This does the inverse of separate()

**Table 5**
```{r}
#Creating tibble from the data in the workbook
table5 <- tibble(
  country = c("Afghanistan", "Afghanistan", "Brazil", "Brazil", "China", "China"),
  century = c(19, 20, 19, 20, 19, 20),
  year = c("99","00","99","00","99","00"), # put them in quatas to keep dubble zero on the 2000
  cases = c(745, 2666, 37737, 80488, 212258, 213766),
  population = c(19987071, 20595360, 172006362, 174504898, 1272915272, 1280428583)
) # This is a tidy datasett
table5
```

```{r}
table5 %>% 
  unite(new, century, year, sep = "")
```

# 2.8 Wrangling strings and dates

**Standardising text with** stringr

```{r}
# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)


# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
        # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean), # i moved nuber 3 up so it removes the invisible space before they become _
    # 2. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_")
  )

print(clean_sites)
clean_sites

```

**Mastering time with** lubridate

```{r}
# Parsing different date formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2

```

Create a tibble with time and tmeprature.
```{r}
# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)

```

# 2.9 Relational data: Joining tables

Data split across multiple tables, here I will learn how to bring two tables together.

**Create 2 tables:**
```{r}
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)

```

**The workhorse:** left_ and right_join()

The left and right jon works the same, it just depend on which table you want to use as a base, the left koin keeps all the rows in the left table and matches data from the right table. 
If matches does not exist it fills gaps with NA

```{r}
# Joining metadata to our observations
joined_data <- observations %>% 
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)

```

**The strict filter:** inner_join()

Inner_join only keeps rows that have complete matches in both tables.

```{r}
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))
print(matched_data)
```

**The detective:** anti_join()
This one filters the left table to only show rows that do not have a match in the right table.

```{r}
# Which observations are missing from our metadata dictionary?
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code))
print(missing_context)
```

# 2.10 Handling missing values

**Converting legacy errors with** na_if()

**Create tibble**
```{r}
# Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

# Convert the -999 error codes to true NA values
fixed_logger <- logger_data %>% 
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)

```

**Replacing missing values with** coalesce()
Sometimes NA does not mean missing data, but it can be a known fixed number, often zero. 

```{r}
# Simulate some count data
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen

# Replace NA with 0 
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)

```

**The special case of** NaN (Not a Number)
NaN means "this value is mathematically impossible"

```{r}
cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))

# Calculating CPUE (catch / effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours) # 0/0 canot be calculated

print(cpue_calc)

```

**Implicit missing values and the zero-data framework:** complete()

```{r}
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8))
raw_catch
```

```{r}
# Force inclusion of hidden zero-catch data
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))

print(full_catch_matrix)

```

**The nuclear option:** drop_na()
If NA reprecent wrong data of failed insturmental readings, and it needs to be removed. drop_na() will remove the entire row
```{r}
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)

# Remove any row where salinity is missing
clean_log <- sensor_log %>% 
  drop_na(salinity)

sensor_log
clean_log
```

# 2.11 Practical Exercises: Penguins, Pivots, and Relational Data

In this exercise, you will combine all the skills you have learned in this workshop. You have the clean penguins dataset, but a colleague has just emailed you a supplementary metadata file containing the GPS coordinates and the date the primary weather stations were installed on each island.

#### Exercise 1: Cleaning the messy metadata

This is the messy data table
```{r}
# The messy metadata provided by your colleague
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

print(island_metadata)

```

**Task:** Create a new object called clean_metadata. Using mutate(), stringr, and lubridate, fix the island_name column so that it perfectly matches the island column in the penguins dataset (hint: watch out for those hidden spaces!). Then, convert the station_install column into a proper Date object.

```{r}
penguins
```



```{r}
# clean the Island names so they match the original penguins data
clean_metadata <- island_metadata %>% 
  # Reanames the column to match the one in the penguins df
  rename(island = island_name) %>% 
  mutate(
    # 1. Convert everything to lowercase
    island  = str_to_sentence(island),
        # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
    island  = str_trim(island), # i moved nuber 3 up so it removes the invisible space before they become _
    # 2. Replace any spaces with underscores
    island = str_replace_all(island, pattern = " ", replacement = "_")
  )
clean_metadata
```


```{r}
# Fix the station_install column into a proper date object
clean_metadata<- clean_metadata %>% 
mutate(
    station_install = dmy(station_install))
```

#### Exercise 2: The Relational Join

**Task:** Create a new object called penguins_spatial. Use a left_join() to attach your clean_metadata to the main penguins dataset. Hint: Because the column containing the island names is called island in the penguins dataset, and island_name in your metadata, you will need to specify how they match using by = join_by(island == island_name).

```{r}
# Joining metadata to our observations
penguins_spatial <- penguins %>% 
  left_join(clean_metadata, by = join_by(island))

```

Check that it worked
```{r}
# Scroll to the right to see the new columns
head(penguins_spatial)
#it did work
```

#### Exercise 3: The Wide Summary Matrix

A collaborator has asked for a clean, publication-ready table showing the maximum body mass recorded for each species, separated by island.

**Task:** Starting with your penguins_spatial dataset:

Drop any missing values in the body_mass_g column.

1.	Group by species and island.

2.	Summarise the data to find the maximum body mass (max()).

3.	Pivot the dataset wider so that the species form the rows, and the islands form the columns.

```{r}
max_body_mass <- penguins_spatial %>% 
  drop_na(body_mass_g) %>% 
  group_by(species, island)%>%
summarise(max_mass=max(body_mass_g),
          .groups = "drop") %>% 
  pivot_wider(
    names_from = island,
    values_from = max_mass
  )

max_body_mass
```


# 2.12 Keystone exercise: Estuary Fish Survey - Data Rescue Mission

**The Premise:** You have inherited a project from a former PhD student who was studying the relationship between water quality and predatory fish assemblages along the Ross River Estuary gradient. Unfortunately, their data management was entirely manual, and the files are a disaster.

They left you with four separate datasets:

1.	estuary_catch_log.xlsx: Handwritten field counts of fish species (using common names), spread across multiple spreadsheet tabs.

2.	estuary_metadata.csv: The spatial coordinates and estuary zones for each site.

3.	estuary_sonde_data.csv: High-frequency water quality sensor readings.

4.	species_dictionary.csv: A taxonomic lookup table linking local common names to their accepted scientific names.

**Your mission** is to write a single, reproducible R script (or .qmd) that ingests this mess, cleans it, joins it into a single master dataset, and produces a publication-ready visualisation of the estuary's ecology.
# ==============================================================================
# Phase 1: Ingestion and Decontamination
# ==============================================================================

**1. Load and bind all tabs from the catch log dynamically**
```{r upload all the tabs}
# The catch log has one tab per site, so I read each tab into its own df
catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = "Lower_ross")
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)

# Have a look at them all, looking for differences or anything weird things
catch1
catch2
catch3
catch4
```

The column names look the same, but the names in the collumns look diffrent, so i can join them together 


**2. String Standardisation for catch log**
```{r Join the catch sheets}
all_catch <- bind_rows(catch1, catch2, catch3, catch4)
# Remove the separate tabs to keep the environment clean
rm(catch1, catch2, catch3, catch4) 
all_catch
```

**Preliminary checks**
```{r inspect-all_catch}
#check the structure: column names, types and the first few rows
all_catch
head(all_catch)
glimpse(all_catch)
str(all_catch)
```
glimpse () dosen't tell me how many sites or species there are, so I need to check this, i use distinct () to check.

```{r distinct-catch}
# distinct () lists each unique value in a column. this gives an overview of the site and species names, and show what i need to standarise later on.
all_catch %>% 
  distinct(site)

all_catch %>%  
  distinct(species)
```


**3. Load and standardise metadata**
```{r metadata}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
metadata
```

**Checks**
```{r inspect-metadata}
# Quick look at the columns and types
head(metadata)
glimpse(metadata)
```
Names are no good in ether of the two (metadata and the catch sheets), this needs to be fixed. 

```{r Tidy names}
# clean Site and species in the catch data:
all_catch_clean <- all_catch %>% 
  mutate(
    # 1. Convert everything to lowercase
    site  = str_to_lower(site),
    species = str_to_lower(species),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site  = str_trim(site),
    species  = str_trim(species),
    # 3. Replace any spaces with underscores
    site = str_replace_all(site, pattern = " ", replacement = "_"),
    species = str_replace_all(species, pattern = " ", replacement = "_"),
    
  ) %>% # Convert all character columns to factor
  mutate(across(where(is.character), as_factor)
  )

all_catch_clean # view the clean data

# Do the same for site_name in metadata
metadata <- metadata %>% 
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) %>% 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
metadata # view the data
```

**Recheck**
```{r recheck-site-names}
# Check the cleaning worked and the site names look right
all_catch_clean %>%  
  str()
all_catch_clean %>% 
   distinct(site)
str(metadata)

# Note: The column is called site in the catch data but site_name in the metadata, so i'll need to tell the join they're the same thing later on
```

**4. Load and clean species dictionary**
```{r load-species-dict}
# The dictonary link the comon names to the scientific names
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
# Check
head(species_dict)
```

Check whether the species names match the common names in the dictionary 
```{r check-species-names}
# Species names in the catch data
all_catch_clean %>%  
  distinct(species)
# Check if they are diffrent from the species dictionary
species_dict %>% 
  distinct(common_name)
```
the names match, the only difference is the column name: "species" in the catch data and "common_name" in the dictionary.


**Create new variable ("scientific_name") in catch log by joining with taxonomic dictionary**
```{r join-species-dict}
all_catch_clean <- all_catch_clean %>% 
  mutate(common_name = species) %>% # Make common_name column (a copy of species) so it has the same name as the column in th dictionary
  left_join(species_dict, by = join_by(common_name)) # Left_join keeps all the catch data and adds the scientific_name on from the dictionary
all_catch_clean # Check it worked 
```

Doubble check it now has 6 columns
```{r}
glimpse(all_catch_clean)
```



**5. Load and clean high-frequency sonde water quality data**
This is already done in the div_sensor
```{r load-sonde}
# This runs the diw_sensor.R script, which cleans the sonde data, and makes sensor_daily: the daily mean temprature, salinity and turbidity for each site
source(here::here("code/diw_sensor.R"))
```
# ==============================================================================
# Phase 2: The Relational Architecture
# ==============================================================================

**The Grand Join: Merge catch, spatial metadata, and daily sensor summaries**
```{r join-all}
# The grand join: starts with the catch data and add on the metadat and daily sensor data
catch_data_final <- all_catch_clean %>% 
  # Site in the catch data is called site_name in metadata so i tell join_by() they match with the dubble equal (==)
  left_join(metadata, by = join_by(site == site_name)) %>% 
  # sensor_daily has site and date columns with the same names, so these already match. 
  left_join(sensor_daily, by = join_by(site, date))
```


**Check**
```{r check-final-join}
# Check the join worked and the new columns are there
glimpse(catch_data_final)
```
looks good, but there are some missing rows

4 x species
4 x sites
30 surveys

4x4x30 = 470

There should be 480 rows not just 427. 

This might be because there were some zero counts of species at different sites. i need to change these so they are actually 0 and NA.

# ==============================================================================
# Phase 3: The Zero-Catch Framework
# ==============================================================================

**Ensure every species is represented at every site, zone, and date combination**
```{r wrangle-final-zeros}
catch_data_final <- 
catch_data_final %>% 
  # complete () adds a row for every species at every site and date, so the days a species was not caught now show up as rows too.
  # complete(nesting()) keeps each site/date with its zone and sensor values as one unit, so complete() only uses real site/dates combinations and not evry posible mix
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity, mean_turbidity),
    # everyone of these gets paired with every species
    scientific_name
  ) %>% 
  mutate(
    # The new rows have NA for counts, so convert those NAs to 0 with coalesce ()
    count = coalesce(count, 0)
  )

catch_data_final
```

Now there are 480 rows, yay
# ==============================================================================
# Phase 4: Statistical Extraction (Summary Table Deliverable)
# ==============================================================================
**Salinity**
Is there a difference in salinity at the diffrent sites?
```{r Salinity-summary}
# Make a table of the median salinity and sd for each zone
sal_sum <- 
  catch_data_final %>% 
  group_by(zone) %>% 
  summarise(
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

# Plot the table: a point for the median with sd error bars
sal_sum %>%  
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
```
Yes there is a difrence in salinity at each site. 

**Turbidity**
Is there a difrence in turbidity at each site?
```{r Turbidity-summary}
# same as salinity: median and sd of turbidity for each zone
turb_sum <- 
  catch_data_final %>% 
  group_by(zone) %>% 
  summarise(
    sd_turbidity = sd(mean_turbidity, na.rm = TRUE),
    median_turbidity = median(mean_turbidity, na.rm = TRUE),
    .groups = "drop"
  )

# plot it the same way as salinity
turb_sum %>% 
  ggplot(aes(x = zone, y = median_turbidity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_turbidity - sd_turbidity, ymax = median_turbidity + sd_turbidity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Turbidity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Turbidity (NTU)")
```

There is some diffrence, but not a lot. 

**Temprature**
Is there any diffrence?
```{r Temprature-summary}
# Same again for temperature
temp_sum <- 
  catch_data_final %>% 
  group_by(zone) %>% 
  summarise(
    sd_temperature = sd(mean_temp, na.rm = TRUE),
    median_temperature = median(mean_temp, na.rm = TRUE),
    .groups = "drop"
  )

temp_sum %>% 
  ggplot(aes(x = zone, y = median_temperature)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_temperature - sd_temperature, ymax = median_temperature + sd_temperature), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Temperature by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Temperature (°C)")

```

There is some diffrence, but not a lot. 

**Generate statistical summary table for fish counts and salinity per species and zone**
```{r}
statistical_summary <- catch_data_final %>% 
  group_by(zone, scientific_name) %>% 
  summarise(
    # Fish Count statistics (Mean ± SEM)
    mean_count = mean(count, na.rm = TRUE),
    se_count = sd(count, na.rm = TRUE) / sqrt(n()),
    
    # Salinity statistics (Mean ± SEM)
    mean_sal = mean(mean_salinity, na.rm = TRUE),
    se_sal = sd(mean_salinity, na.rm = TRUE) / sqrt(n()),
    
    n_observations = n(),
    .groups = "drop"
  ) %>% 
  # Combine into clean "Mean ± SE" table
  mutate(
    fish_catch_summary = paste0(round(mean_count, 2), " ± ", round(se_count, 2)),
    salinity_summary = paste0(round(mean_sal, 2), " ± ", round(se_sal, 2))
  )

# View the final summary table
statistical_summary %>% 
  select(zone, scientific_name, n_observations, fish_catch_summary, salinity_summary) %>% 
  knitr::kable(
    col.names = c("Estuary Zone", "Scientific Name", "N", "Fish Catch (Mean ± SE)", "Salinity (ppt Mean ± SE)"),
    align = c("l", "l", "c", "c", "c"),
    caption = "Table 1. Summary of fish abundance and water salinity across estuarine zones."
  )
```

# ==============================================================================
# Phase 5: Visual Communication (Publication-Ready Multi-Panel Plot)
# ==============================================================================

**Publication ready salinity plot**
```{r Salinity-plot}
#| fig-cap: "Salinity Gradient across zones. Ross River Estuary System, Townsville, NQ, Australia May 2026."

# same plot as the salinity one above, but with a figure caption added
sal_sum %>%  
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
  
```

#### Fish abundance patterns

**Calculate summary stats**
```{r Fish-stats}
# mean catch and standard error (SEM) for each species in each zone
fish_stats <- catch_data_final %>% 
  group_by(zone, scientific_name) %>% 
  summarise(
    mean_catch = mean(count), # Calculates mean 
    sem_catch = sd(count) / sqrt(n()), # Calculates SE of Mean SEM
    .groups = "drop"
  )

fish_stats
```

**Plot the fish data**
```{r Fish-plot}
fish_stats %>%  
  ggplot(aes(x = zone, y = mean_catch, fill = scientific_name)) +
  geom_line(group = 1) + # line joining the zone means
  # add the raw catch counts from the master data as faint jittered points
  # inherit.aes = FALSE so these use the master data's columns, not fish_stats
  geom_point(data = catch_data_final, aes(x = zone, y = count),
             position = position_jitter(width = 0.2), alpha = 0.2, size = 1, inherit.aes = FALSE) + 
# diamonds for the mean catch (pch 23 is the shape that takes a fill colour)
  geom_point(position = "dodge", size = 4, pch = 23) +
  # SE pretty narrow, potentially not worth plotting but let's viewer see them
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) + 
  # one panel for each species
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"), # Species names in italic
   # legend.position = "bottom"
    # no legend needed because the scientific names are above each panel
    legend.position = "none"
  ) +
  labs(
    title = "Predatory Fish Abundance across Estuarine Salinity Gradients", # default ggplot title appears on top
    subtitle = "Ross River Estuary System, May 2026",
    x = "Spatial Zone",
    # scientific label for y axis of n day -1 use regex to get superscript
    y = expression("Abundance (n day"^-1*")"),
  # expression() gives me the superscript for n day^-1
    y = "Abundance (per day)"
  )
```


