---
title: "Workshop 1: Foundations of Marine Data Science"
author: Celine M. Egsaeter
format:
  html:
    embed-resources: true
execute:
  warning: false
  message: false
---


```{r}
#| label: Housekeeping
## Inventory every active object currently residing in session RAM
objects()
## Purge global environment
rm(list = ls())
# Confirm that global session memory now completely vacant
objects()
```

# 1.4 Ingesting data into R

```{r}
#| label: Library
# Load the primary data science framework and Excel import library
library(tidyverse)
library(readxl)
```

```{r}
#use the package "here" to specify in which of the folders the data sett is in under the rproj

# Practice Import A: Loading a standard comma-separated plain text file
benthic_cover <-read_csv(here::here("data/workshop1/reef_cover_log.csv"))
benthic_cover #cheking it looks good 

# Practice Import B: Parsing a tab-separated telemetry instrument array string
acoustic_stream <- read_tsv(here::here("data/workshop1/acoustic_telemetry_stream.txt"))
acoustic_stream

# Practice Import C: Targeting a specific sheet in a multi-tab Excel spreadsheet
fisheries_annual <- read_excel(here::here("data/workshop1/fish_catch_data.xlsx"), sheet ="Commercial_2026")
fisheries_annual
```


```{r}
# Read in mangrove_data
#mangrove_data <- read_csv(file = here::here("data/workshop1/mangrove_survey_raw.csv"))
# This method creates a messy corupt data
```

```{r}
# Use args within read_csv to skip headers and declare missing flags
mangrove_data <- read_csv(
  here::here("data/workshop1/mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA

mangrove_data # Have a look at the file 
```

# 1.5 Data frame architectures: Tibbles versus legacy tables

```{r}
# Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover_df <- as.data.frame(benthic_cover)
# Print the old-style dataframe structure to view
print(benthic_cover_df)
# And compare with tibble alternative
print(benthic_cover) # The tibble tells me if the collumn is a chr, dbl, or a fct
```

# 1.6 Wrangling out ecological signals using Palmer Penguins

```{r}
# Load the package data into active memory
library(palmerpenguins)
data("penguins")

# Examine the structure of the dataset - always do this when loading a new dataset!
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version

```

```{r}
# Generate an exploratory summary matrix
summary(penguins)
```

# 1.7 Foundational grammar: Slicing, filtering, sorting, and transforming

## 1.7.1 Isolating attributes with select()

```{r}
# Vertically slice specific morphometric variables by explicit name
morphology_metrics <- select(penguins,species, bill_length_mm,bill_depth_mm,body_mass_g)
glimpse(morphology_metrics)

# Retain a continuous block of attributes using the colon operator
spatial_block <- select(penguins, species:island)
glimpse(spatial_block)

# Discard logistics tracking attributes while preserving everything else using the minus sign
clean_scientific_fields <- select(penguins, -year)
glimpse(clean_scientific_fields)
```

```{r}
# Isolate observations belonging to a single categorical target group
adelie_cohort <- filter(penguins, species == "Adelie")
glimpse(adelie_cohort)

# Sift out individuals using continuous numerical boundary thresholds
# Preserves only large penguins whose mass exceeds 4500 grams
heavy_penguins <- filter(penguins, body_mass_g > 4500)

# Combine multiple conditional parameters across separate attributes
# Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")

# Sift records matching multiple targeting flags within an explicit set
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))

```

## 1.7.3 Ordering sequences with arrange()
```{r}
# Sort penguins by ascending body mass (Default setting: Smallest mass first)
lightest_first <- arrange(penguins, body_mass_g)
glimpse(lightest_first)

# Sort penguins in descending sequence using the desc() layout wrapper
heaviest_first <- arrange(penguins, desc(body_mass_g))

# Execute nested sorting criteria: Group by species, then sort by descending bill length
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
```

## 1.7.4 Introducing the Pipe (|>)

Example: Instead of:

```{r}
#penguins_subset <- mutate(penguins, bill_ratio = bill_length_mm / bill_depth_mm)
#penguins_final <- filter(penguins_subset, species == "Adelie")
```

You can simply write:
```{r}
penguins_final <- penguins %>% 
  mutate(bill_ratio=bill_length_mm/bill_depth_mm) %>% 
  filter(species=="Adelie")
penguins_final
```




## 1.7.5 Computing new attributes with mutate()

```{r}
# Calculate a new morphological ratio in our environment
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )

# View your newly engineered variables appended to the far-right columns
glimpse(penguin_ratios)

```

# 1.8 Data aggregation and ecological summarisation

```{r}

# Grouping our active memory penguins by species
grouped_penguins <- group_by(penguins, species)

# Notice that the table looks identical, but metadata notes 'Groups: species [3]'
print(grouped_penguins)

# Collapsing the buckets into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)


```

```{r}
# Overcoming the missing value trap using na.rm = TRUE
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),# Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE), # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE) # Standard deviation calculation
  )

print(biological_signal)

```

# 1.9 Integrating data grammar with visual diagnostics in qmd

```{r}
# Pipe directly from aggregation to plotting with error bars
mass_compare_plot <-  penguins %>% 
  group_by(species, island) %>% 
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) + #position helps move the point of the penguin species more visble
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2, position = position_dodge(width = 0.5)) + # position helps move the error bars with the pints, making all the island and species visable.
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot

```

**Pause for thought: Think about what this figure is telling you about the biology of these species of penguins on these different islands. What do you think a statistical test would tell you here?**

**Answer:** 
Gentoo penguins, the heaviest species were found only on Biscoe Island, and Chinstrap penguins were found only on Dream Island. Adelie penguins were the only species found on all three islands,and their mean body mass looks similar across the three error bars. This suggest that island is not the main driver of body size, and that body size is mostly a species trait.

To test this, i would compare the body mass of Adelie penguins between the three islands. if the test found no significant difference, that would mean i had found no evidence that island affects Adelie body mass.
 

**Challenge:**
**In your .qmd template, try creating a new plot. Experiment with swapping sd_mass (standard deviation) for standard error—you will need to divide your sd by the square root of n (sd / sqrt(n)). See how that changes your visual interpretation of the data uncertainty. Present the summary table alongside the associated figure and write down your thoughts on the underlying ecological story that the data are telling.**


```{r}
penguins_SE <- penguins %>%
group_by(species, island) %>%
  summarise( n = n(),
             mean_mass = mean(body_mass_g, na.rm = TRUE),
             error_mass = sd(body_mass_g, na.rm = TRUE) / sqrt(n()),
             .groups = "drop")

knitr::kable(penguins_SE, digits = 1)
```


```{r}
penguins_SE %>% 
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3,position = position_dodge(width = 0.5)) + #position helps move the point of the penguin species more visble
  geom_errorbar(aes(ymin = mean_mass - error_mass,
                    ymax = mean_mass + error_mass),
                width = 0.2, position = position_dodge(width = 0.5)) + # position helps move the error bars with the pints, making all the island and species visable.
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent Standard Error",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()
```


- they all got a lot shorter...
