---
title: "Workshop 2: Advanced data wrangling: Extracting ecological signals from noisy systems"
author: "Celine M. Egsaeter"
format:
  html:
    embed-resources: true
execute:
  warning: false
  message: false
---

#### Housekeeping
```{r}
## Inventory every active object currently residing in session RAM
objects()
## Purge global environment
rm(list = ls())
# Confirm that global session memory now completely vacant
objects()
```

#### Library
```{r}
# Load the primary data science framework
library(tidyverse)
library(palmerpenguins)
library(lubridate)
```


# 2.2 Tidying data

1.	Put each dataset in a tibble (special dataframe)

**table 1**
```{r}
#Creating tibble from the data in the workbook
table1 <- tibble(
  country = c("Afghanistan", "Afghanistan", "Brazil", "Brazil", "China", "China"),
  year = c(1999, 2000, 1999, 2000, 1999, 2000),
  cases = c(745, 2666, 37737, 80488, 212258, 213766),
  population = c(19987071, 20595360, 172006362, 174504898, 1272915272, 1280428583)
) # This is a tidy datasett
table1
```

**table 2**
```{r}
table2 <- tibble(
  country = c("Afghanistan", "Afghanistan","Afghanistan","Afghanistan", "Brazil", "Brazil", "Brazil", "Brazil", "China", "China","China", "China"),
year = c(1999,1999,2000, 2000,1999,1999,2000, 2000,1999,1999,2000, 2000),
type =  c("cases","population","cases","population", "cases","population","cases","population","cases","population","cases","population"),
count = c(745, 19987071, 2666,20595360,37737,172006362,80488,174504898,212258,1272915272,213766,1280428583)) # This data is the same but is messy
table2
```

**table 3**
```{r} 
table3 <- tibble(
  country = c("Afghanistan", "Afghanistan", "Brazil", "Brazil", "China", "China"),
  year = c(1999, 2000, 1999, 2000, 1999, 2000),
  rate = c("745/19987071", "2666/20595360", "37737/172006362",
           "80488/174504898", "212258/1272915272", "213766/1280428583")
)
table3
```


```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases)
```


```{r}
# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

```

# 2.3 Pivoting data to make it tidy

# 2.4 Lengthening datasets

```{r}
# We are using the data billboard as practise
billboard
```
**pivot longer**
```{r}
billboard %>% pivot_longer( # Pivot longer creates a long df of a wide one.
  cols = starts_with("wk"), # chooses the collums starting with wk
  names_to = "week", # puts the names from wk, into a collumn named week
  values_to = "rank", # adds the values to a coloumn named rank
  values_drop_na = TRUE
)
```

**Pivoting longer - example**
```{r}
df <- tribble( # Save the tribble as a dataset called df
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)
df
```

However, we want our new (tidy) dataset to have three variables: 
1.	id (which already exists)
2.	measurement (the column names) 
3.	value (the cell values)

**Pivot longer**
we use pivot longer to get 3 variables
```{r}
df %>% 
  pivot_longer(
    cols = bp1:bp2, # The chosen columns to that we are going to pivot into rows 
    names_to = "measurement", # The names is the bp1 and bp2 theire collumn is going to be called measurment 
    values_to = "value" # The values they hold will go into a difrent collumn called value
  )

```

# 2.5 Widening datasets

Lengthening the dataset is the oposite of widening it, and we use the function pivot_wider

For this i will use an example from R4DS, the cms_patient_experience dataset 

```{r}
cms_patient_experience %>% 
  distinct(measure_cd, measure_title) # shows the complete sett of values of the df
```

**Pivot wider**
```{r}
cms_patient_experience %>% 
    pivot_wider(
    id_cols = starts_with("org"), # Says which columns identify a row, and pics every column that starts with "org"
    names_from = measure_cd, # says whcih column values become the new column names
    values_from = prf_rate # says which column supplies the numbers that fill the new columns.
  )
```

Adding the "id_cols = starts_with("org")" line helped us make the output we wanted, not adding this line made the df output look a bit off. 

**Pivoting wider - example**
Changed df to df_2 so it is not the same as the earlier example.

```{r}
# Creating the df
df_2 <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```


```{r}
df_2 %>% 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```

```{r}
df_2 %>% 
  distinct(measurement) |> 
  pull() #chek the unique values
```
```{r}
df_2 %>% 
  select(-measurement, -value) |> 
  distinct()
```

pivot_wider can now combie these results to generate an empty df

```{r}
df_2 %>% 
  select(-measurement, -value) %>% 
  distinct() %>%  
  mutate(x = NA, y = NA, z = NA) 
# creates a empty df
```


# 2.6 Pivoting Exercises using Palmer Penguins

**Lengthening data with** pivot_longer()

```{r}
#Create a long df version of the penguins df
penguins_long <- penguins %>% 
  pivot_longer(
    cols = c(bill_length_mm,bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
# view the result
head(penguins_long) 
```

**Create ggplot**
```{r}
penguins_long %>% 
  drop_na(value) %>% # Remover the NA values in the value column
    ggplot(aes(x = value, fill = species)) + # Tells ggplot what values go where
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") + 
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )

```

**Widening data with** pivot_wider()
Time to do the oposite

```{r}
# Calculate the mean body mass for each species
mass_summary <- penguins %>% 
  drop_na(body_mass_g) %>% 
  group_by(species, island) %>% 
  summarise(mean_mass = mean(body_mass_g))
# View the df
head(mass_summary)

```
The above verion is tidy but not great for publication

```{r}
# Creating a clean matrix of species by Isalnd
mass_matrix <- mass_summary %>% 
  pivot_wider( #Using pivot_wider to crate a column for each island
    names_from = island,
    values_from = mean_mass
  )
# view the df
head(mass_matrix)

```


# 2.7 Separating and uniting data tables

Using the function separete() to tidy table 3 from earlier

**Here’s table3:**
```{r}
head(table3)
```

I want to split the rate column up into two variables 
1. cases 
2. Population

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))

# same code but uses the "sep=" to specify what charakter to separate into two coloumns

table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")

# In both these something went wrong, i am asuming it is beceause both tibbles have population and cases in the chr class
```

**Note** Problem above i needed to change the way i created the tibble at the begining, the rate had to be written in quatas, sutch as "745/19987071", if not R does the calculation and removes the "/". 

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE) # Convets = TRUE keeps the two new columns as int

```

**Using** unite():
This does the inverse of separate()

**Table 5**
```{r}
#Creating tibble from the data in the workbook
table5 <- tibble(
  country = c("Afghanistan", "Afghanistan", "Brazil", "Brazil", "China", "China"),
  century = c(19, 20, 19, 20, 19, 20),
  year = c("99","00","99","00","99","00"), # put them in quatas to keep dubble zero on the 2000
  cases = c(745, 2666, 37737, 80488, 212258, 213766),
  population = c(19987071, 20595360, 172006362, 174504898, 1272915272, 1280428583)
) # This is a tidy datasett
table5
```

```{r}
table5 %>% 
  unite(new, century, year, sep = "")
```

# 2.8 Wrangling strings and dates

**Standardising text with** stringr

```{r}
# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)


# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
        # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean), # i moved nuber 3 up so it removes the invisible space before they become _
    # 2. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_")
  )

print(clean_sites)
clean_sites

```

**Mastering time with** lubridate

```{r}
# Parsing different date formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2

```

Create a tibble with time and tmeprature.
```{r}
# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)

```

# 2.9 Relational data: Joining tables

Data split across multiple tables, here I will learn how to bring two tables together.

**Create 2 tables:**
```{r}
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)

```

**The workhorse:** left_ and right_join()

The left and right jon works the same, it just depend on which table you want to use as a base, the left koin keeps all the rows in the left table and matches data from the right table. 
If matches does not exist it fills gaps with NA

```{r}
# Joining metadata to our observations
joined_data <- observations %>% 
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)

```

**The strict filter:** inner_join()

Inner_join only keeps rows that have complete matches in both tables.

```{r}
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))
print(matched_data)
```

**The detective:** anti_join()
This one filters the left table to only show rows that do not have a match in the right table.

```{r}
# Which observations are missing from our metadata dictionary?
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code))
print(missing_context)
```

# 2.10 Handling missing values

**Converting legacy errors with** na_if()

**Create tibble**
```{r}
# Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

# Convert the -999 error codes to true NA values
fixed_logger <- logger_data %>% 
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)

```

**Replacing missing values with** coalesce()
Sometimes NA does not mean missing data, but it can be a known fixed number, often zero. 

```{r}
# Simulate some count data
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen

# Replace NA with 0 
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)

```

**The special case of** NaN (Not a Number)
NaN means "this value is mathematically impossible"

```{r}
cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))

# Calculating CPUE (catch / effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours) # 0/0 canot be calculated

print(cpue_calc)

```

**Implicit missing values and the zero-data framework:** complete()

```{r}
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8))
raw_catch
```

```{r}
# Force inclusion of hidden zero-catch data
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))

print(full_catch_matrix)

```

**The nuclear option:** drop_na()
If NA reprecent wrong data of failed insturmental readings, and it needs to be removed. drop_na() will remove the entire row
```{r}
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)

# Remove any row where salinity is missing
clean_log <- sensor_log %>% 
  drop_na(salinity)

sensor_log
clean_log
```

# 2.11 Practical Exercises: Penguins, Pivots, and Relational Data

In this exercise, you will combine all the skills you have learned in this workshop. You have the clean penguins dataset, but a colleague has just emailed you a supplementary metadata file containing the GPS coordinates and the date the primary weather stations were installed on each island.

#### Exercise 1: Cleaning the messy metadata

This is the messy data table
```{r}
# The messy metadata provided by your colleague
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

print(island_metadata)

```

**Task:** Create a new object called clean_metadata. Using mutate(), stringr, and lubridate, fix the island_name column so that it perfectly matches the island column in the penguins dataset (hint: watch out for those hidden spaces!). Then, convert the station_install column into a proper Date object.

```{r}
penguins
```


```{r}
# clean the Island names so they match the original penguins data
clean_metadata <- island_metadata %>% 
  mutate(
    # 1. Convert everything to lowercase
    island_name  = str_to_sentence(island_name),
        # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
    island_name  = str_trim(island_name ), # i moved nuber 3 up so it removes the invisible space before they become _
    # 2. Replace any spaces with underscores
    island_name  = str_replace_all(island_name, pattern = " ", replacement = "_")
  )
clean_metadata
```


```{r}
# Fix the station_install column into a proper date object
clean_metadata<- clean_metadata %>% 
mutate(
    station_install = dmy(station_install))
```

#### Exercise 2: The Relational Join

**Task:** Create a new object called penguins_spatial. Use a left_join() to attach your clean_metadata to the main penguins dataset. Hint: Because the column containing the island names is called island in the penguins dataset, and island_name in your metadata, you will need to specify how they match using by = join_by(island == island_name).

```{r}
# Joining metadata to our observations
joined_data <- observations %>% 
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)

```

